Linear stability of magnetized massive protoplanetary disks

### 2014-04-05 paper/appendix.tex
first draft, spell-checked

\begin{figure}
  \includegraphics[width=\linewidth]{figures/plot_fq}
  \caption{The function $f(Q)$ describing vertical hydrodstatic 
    equilibrium in self-gravitating polytropic disks (solid line). The
    horizontal dashed line is the asymptotic value of $1/\sqrt{2}$ for
    large $Q$.  
    \label{plot_fq}}
\end{figure}

@@

\subsection{Tilted field with uniform resistivity}
Here we allow $\epsilon\neq 0$ but take $\eta$ to be constant. We
first obtain expressions for $\dbx$ and $\dby$. Differentiating
the $x$ momentum equation and replacing the resuting $\dbz^\prime$
using the divergence-free condition and $\dby^\prime$ using the $y$
momentum equation, we obtain an expression for $\dbx^{\prime\prime}$
which can be inserted into the $x$ induction equation. This gives
\begin{align}
  \imgi\sigma\dbx = B_z\dvx^\prime +
  \frac{\eta\mu_0\rho}{B_z}\left( D_1 U + \imgi\epsilon k_x D_0
  V\right).\label{bx_tilted}  
\end{align}
We can insert this into the $y$ induction equation to obtain
\begin{align}
  \imgi\sbar\dby = -B_y\Delta + B_z\dvy^\prime + \frac{\imgi
    S}{\sigma}\left[ B_z \dvx^\prime+ \frac{\eta\mu_0\rho}{B_z}\left( D_1 U + \imgi\epsilon k_x D_0
  V\right)\right] + \frac{\eta\mu_0\rho}{B_z}D_1V,\label{by_tilted}
\end{align} 
where we have also used the derivative of the $y$ momentum equation to
eliminate  $\dby^{\prime\prime}$. Recall $\Delta \equiv i k_x \dvx +
\dvz^\prime$, so that
\begin{align}
  \Delta = \imgi k_x \dvx + \frac{i}{\sigma}\left(\w^{\prime\prime} +
  \epsilon V^\prime\right) = -\left[\frac{\imgi\sigma W}{c_s^2} +
    \frac{\imgi\left(\ln{\rho}\right)^\prime}{\sigma}\left(\w^\prime +
    \epsilon V\right)\right], 
\end{align}
where the second equality results from the continuity equation.

>>>>

\begin{figure}
  \includegraphics[width=\linewidth]{figures/plot_fq}
  \caption{The function $f(Q)$ describing vertical hydrostatic 
    equilibrium in self-gravitating polytropic disks (solid line). The
    horizontal dashed line is the asymptotic value of $1/\sqrt{2}$ for
    large $Q$.  
    \label{plot_fq}}
\end{figure}

@@

\subsection{Tilted field with uniform resistivity}
Here we allow $\epsilon\neq 0$ but take $\eta$ to be constant. We
first obtain expressions for $\dbx$ and $\dby$. Differentiating
the $x$ momentum equation and replacing the resulting $\dbz^\prime$
using the divergence-free condition and $\dby^\prime$ using the $y$
momentum equation, we obtain an expression for $\dbx^{\prime\prime}$
which can be inserted into the $x$ induction equation. This gives
\begin{align}
  \imgi\sigma\dbx = B_z\dvx^\prime +
  \frac{\eta\mu_0\rho}{B_z}\left( D_1 U + \imgi\epsilon k_x D_0
  V\right).\label{bx_tilted}  
\end{align}
We can insert this into the $y$ induction equation to obtain
\begin{align}
  \imgi\sbar\dby = -B_y\Delta + B_z\dvy^\prime + \frac{\imgi
    S}{\sigma}\left[ B_z \dvx^\prime+ \frac{\eta\mu_0\rho}{B_z}\left( D_1 U + \imgi\epsilon k_x D_0
  V\right)\right] + \frac{\eta\mu_0\rho}{B_z}D_1V,\label{by_tilted}
\end{align} 
where we have also used the derivative of the $y$ momentum equation to
eliminate  $\dby^{\prime\prime}$. Recall $\Delta \equiv i k_x \dvx +
\dvz^\prime$, so that
\begin{align}
  \Delta = \imgi k_x \dvx + \frac{i}{\sigma}\left(\w^{\prime\prime} +
  \epsilon V^\prime\right) = -\left[\frac{\imgi\sigma W}{c_s^2} +
    \frac{\imgi\left(\ln{\rho}\right)^\prime}{\sigma}\left(\w^\prime +
    \epsilon V\right)\right], 
\end{align}
where the second equality results from the continuity equation.

### 2014-04-05 paper/paper.tex
first draft, spell-checked

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity (SG) can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial lengthscales are stabilized. In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial lengthscales on the order of the disk thickness, SG
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

@@

PPDs subject to both MRI and GI are often modeled through separate
turbulent viscosities in a hydrodynamical framework 
\citep{terquem08}. This implicitly assumes that the development of MRI 
and GI can be assessed independently. 
Circumstellar disk models that explicitly combine the equations of magneto-hydrodynamics
and self-gravity have been limited to a few early simulations 
\citep{fromang04,fromang04b,fromang05}. It will be neccessary to 
revisit and extend these pioneering calculations to fully explore the
impact of MRI and GI on the structure and evolution of PPDs. In
preparation of this, it is important to have a thorough understanding
of the stability properties of such systems.

Since compressibility is not fundamental for the MRI, much of the
early stability calculations assume incompressible perturbations
\citep{goodman94,jin96}. However, recent works indicate
compressibility may be important under certain conditions,
such as strong fields \citep{kim00, pessah05,bonanno07}.   
Previous MRI studies have also focused on modes with vanishing
radial wavenumber, because they are the most unstable
\citep{sano99,reyes01}. Self-gravity has minimal effect on such
perturbations in a rotating disk. However, modes
with radial lengthscales on the order of the disk scale height may be
subject to self-gravity. It is therefore of interest to generalize the MRI with
non-zero radial waveumbers to massive disks.

@@

This work marks the beginning of our study of magnetized,
self-gravitating PPDs. We start from linear calculations, which  
have the advantage that a wide range of parameters can be studied at
negligible computational cost. This  allows us to indentify
conditions, if any, under which MRI and GI 
cannot be considered independent. It is also important to have such 
calculations to benchmark and guide future non-linear
simulations.

This paper is organized as follows. \S\ref{setup} lists the governing
equations and describes the disk equilibra under consideration. 
The linear problem is formulated in \S\ref{linear}. The impact
of self-gravity on the MRI with a vertical field is discussed in
\S\ref{result1}, gravitationally unstable disks are considered in
\S\ref{result2}, and equilibra including an azimuthal field is explored in
\S\ref{result3}. We summarize results in \S\ref{summary} with a 
discussion of important extensions to our current models.

>>>>

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity (SG) can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial length scales are stabilized. In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial length scales on the order of the disk thickness, SG
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

@@

PPDs subject to both MRI and GI are often modeled through separate
turbulent viscosity coefficients in a hydrodynamical framework 
\citep{terquem08}. This implicitly assumes that the development of MRI 
and GI can be assessed independently. 
Circumstellar disk models that explicitly combine the equations of magneto-hydrodynamics
and self-gravity have been limited to a few early simulations 
\citep{fromang04,fromang04b,fromang05}. It will be necessary to 
revisit and extend these pioneering calculations to fully explore the
impact of MRI and GI on the structure and evolution of PPDs. In
preparation of this, it is important to have a thorough understanding
of the stability properties of such systems.

Since compressibility is not fundamental for the MRI, much of the
early stability calculations assume incompressible perturbations
\citep{goodman94,jin96}. However, recent works indicate
compressibility may be important under certain conditions,
such as strong fields \citep{kim00, pessah05,bonanno07}.   
Previous MRI studies have also focused on modes with vanishing
radial wavenumber, because they are the most unstable
\citep{sano99,reyes01}. Self-gravity has minimal effect on such
perturbations in a rotating disk. However, modes
with radial length scales on the order of the disk scale height may be
subject to self-gravity. It is therefore of interest to generalize the MRI with
non-zero radial wavenumbers to massive disks.

@@

This work marks the beginning of our study of magnetized,
self-gravitating PPDs. We start from linear calculations, which  
have the advantage that a wide range of parameters can be studied at
negligible computational cost. This  allows us to identify
conditions, if any, under which MRI and GI 
cannot be considered independent. It is also important to have such 
calculations to benchmark and guide future non-linear
simulations.

This paper is organized as follows. \S\ref{setup} lists the governing
equations and describes the disk equilibria under consideration. 
The linear problem is formulated in \S\ref{linear}. The impact
of self-gravity on the MRI with a vertical field is discussed in
\S\ref{result1}, gravitationally unstable disks are considered in
\S\ref{result2}, and equilibria including an azimuthal field is explored in
\S\ref{result3}. We summarize results in \S\ref{summary} with a 
discussion of important extensions to our current models.

### 2014-04-05 paper/results.tex
first draft, spell-checked

\section{MRI in self-gravitating disks}\label{result1}
In this section we focus on the MRI and use the vertical field setup
of case 1. We first consider MRI modes with negligble density/potential
perturbations to see the effect of self-gravity on the MRI through the
background stratification, then go on to examine MRI modes with
density/potential perturbations in massive disks.

@@

Self-gravity appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not surpressed since
Eq. \ref{lambda_ideal}---\ref{lambda_resis} is only exact for
unstratified disks.) Although the function 
$f(Q)$ does not change significantly for the range of $Q$ considered,
the dependence of $\lambda_\mathrm{resis}$ on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. Modes in
Fig. \ref{lambda_poly_resis} have no nodes in the magnetic energy
$E_m$ except at $z\simeq0,\,H$, i.e. only the longest wavelength survives
with large resistivity. 
%self-gravity tips it over

@@

\subsubsection{Layered 
  resistivity} 
Here we consider disks with midplane Elsasser number $\Lambda_0=0.1$
and a variable resistivity profile with
$A=10^2$. Fig. \ref{poly_layer} compares the magnetic  
energies for $Q=0.2,\,1$ and $4$. They have similar growth rates, $\gamma/\Omega
= 0.53,\,0.64$ and $0.66$, respectively. In the non-self-gravitating
limit ($Q=4$), the MRI is effectively surpressed for
$z\lesssim0.5H$. This is consistent with the picture of layered
accretion proposed for non-self-gravitating disks \citep{gammie96,fleming03}. 
However,in the massive disk ($Q=0.2$) the mode occupies a wider vertical
extent because its wavelength (in units of $H$) is larger. This
suggests that in massive disks, the MRI is not well localized to a
sub-layer within the height, even when the resisitvity has a layered
structure.

@@

\subsubsection{Dependence on $k_x$}
The above experiments show that with increasing 
disk self-gravity, the MRI becomes more global in the vertical
direction. We find a similar result in the horizontal direction. 
Fig. \ref{compare_growth_poly_kx} show MRI growth rates as a
function of $k_x$ for a range of $Q$ values. Increasing self-gravity
decreases the cut-off radial wavenumber for the MRI. We checked
that these modes have negligible density perturbations. Then we can 
understand this result by invoking the instability criteria for  
incompressible MRI in an unstratified Keplerian disk,
\begin{align}
  v_A^2(k_z^2 + k_x^2) < 3\Omega^2,
\end{align}
where $k_z$ is a vertical wavenumber \citep{kim00}. Setting $k_z^2\sim
\Omega^2/v_A^2$ and non-dimensionalizing, we find
\begin{align} 
  k_xH \lesssim \frac{\sqrt{\beta}}{f},
\end{align}
where order-unity factors have been dropped. Despite a simplistic
approach, this demonstrates that with increasing self-gravity
(increasing $f$), we expect MRI modes with small radial
lengthscales to be suppressed.

@@

\subsubsection{Resistive disks}
We repeat the above calculation for resistive disks, but fix
$\beta=100$ and vary the midplane Elasser number $\Lambda_0$. Growth
rates are shown in Fig. \ref{gravity_energy_resis}. 
Interestingly, the highly resistive case $\Lambda_0=0.1$ has
comparable magnetic and gravitational energies: at $k_xH\simeq1.3$ we
find $\tau\sim 0.3$, which corresponds to $\avg{E_g}\sim 0.5\avg{E_m}$.  
Fig. \ref{mri_massive_cowling} compares the magnetic
energy of this mode to that computed in the \emph{Cowling  
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally letting $Q\to\infty$ in Eq. \ref{lin_poisson}).  
The growth rate increases when self-gravity is included in the linear
response, since self-gravity is usually destabilizing. However, 
$\gamma$ and $E_m$ are very similar, indicating that the instability
in the self-gravitating calculation is fundamentally still the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{
    Growth rates of MRI modes in an isothermal self-gravitating
    disk with $Q=0.2$ ($Q_\mathrm{2D}=0.72$) at fixed $\beta=100$, 
    for a range of midplane Elasser numbers. The resistivity is
    uniform. The colorbar measures the importance of self-gravity by $\tau$. 
    \label{gravity_energy_resis}}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_resis_sg}
  \caption{Example of a resistive MRI mode with significant
    gravitational potential perturbation. The disk is isothermal. 
    The vertical line
    inidicates $\Lambda=1$.  
    \label{mri_massive_resis}}
\end{figure}

@@

 However, a strong field and/or large resistivity increases the MRI
 vertical wavelength. When the vertical scale of the MRI becomes
 comparable to the disk thickness, i.e. $\lambda\sim H$, the
 perturbed mass across the disk height can contribute to a net potential
 perturbation. We therefore expect a neccessary condition for
 self-gravity to affect the MRI is for the latter to be weak. 
%we are not considering GI yet

>>>>

\section{MRI in self-gravitating disks}\label{result1}
In this section we focus on the MRI and use the vertical field setup
of case 1. We first consider MRI modes with negligible density/potential
perturbations to see the effect of self-gravity on the MRI through the
background stratification, then go on to examine MRI modes with
density/potential perturbations in massive disks.

@@

Self-gravity appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not suppressed since
Eq. \ref{lambda_ideal}---\ref{lambda_resis} is only exact for
unstratified disks.) Although the function 
$f(Q)$ does not change significantly for the range of $Q$ considered,
the dependence of $\lambda_\mathrm{resis}$ on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. Modes in
Fig. \ref{lambda_poly_resis} have no nodes in the magnetic energy
$E_m$ except at $z\simeq0,\,H$, i.e. only the longest wavelength survives
with large resistivity. 
%self-gravity tips it over

@@

\subsubsection{Layered 
  resistivity} 
Here we consider disks with midplane Elsasser number $\Lambda_0=0.1$
and a variable resistivity profile with
$A=10^2$. Fig. \ref{poly_layer} compares the magnetic  
energies for $Q=0.2,\,1$ and $4$. They have similar growth rates, $\gamma/\Omega
= 0.53,\,0.64$ and $0.66$, respectively. In the non-self-gravitating
limit ($Q=4$), the MRI is effectively suppressed for
$z\lesssim0.5H$. This is consistent with the picture of layered
accretion proposed for non-self-gravitating disks \citep{gammie96,fleming03}. 
However,in the massive disk ($Q=0.2$) the mode occupies a wider vertical
extent because its wavelength (in units of $H$) is larger. This
suggests that in massive disks, the MRI is not well localized to a
sub-layer within the height, even when the resistivity has a layered
structure.

@@

\subsubsection{Dependence on $k_x$}
The above experiments show that with increasing 
disk self-gravity, the MRI becomes more global in the vertical
direction. We find a similar result in the horizontal direction. 
Fig. \ref{compare_growth_poly_kx} show MRI growth rates as a
function of $k_x$ for a range of $Q$ values. Increasing self-gravity
decreases the cut-off radial wavenumber for the MRI. We checked
that these modes have negligible density perturbations. Then we can 
understand this result by invoking the instability criteria for  
incompressible MRI in an unstratified Keplerian disk,
\begin{align}
  v_A^2(k_z^2 + k_x^2) < 3\Omega^2,
\end{align}
where $k_z$ is a vertical wavenumber \citep{kim00}. Setting $k_z^2\sim
\Omega^2/v_A^2$ and non-dimensionalizing, we find
\begin{align} 
  k_xH \lesssim \frac{\sqrt{\beta}}{f},
\end{align}
where order-unity factors have been dropped. Despite a simplistic
approach, this demonstrates that with increasing self-gravity
(increasing $f$), we expect MRI modes with small radial
length scales to be suppressed.

@@

\subsubsection{Resistive disks}
We repeat the above calculation for resistive disks, but fix
$\beta=100$ and vary the midplane Elsasser number $\Lambda_0$. Growth
rates are shown in Fig. \ref{gravity_energy_resis}. 
Interestingly, the highly resistive case $\Lambda_0=0.1$ has
comparable magnetic and gravitational energies: at $k_xH\simeq1.3$ we
find $\tau\sim 0.3$, which corresponds to $\avg{E_g}\sim 0.5\avg{E_m}$.  
Fig. \ref{mri_massive_cowling} compares the magnetic
energy of this mode to that computed in the \emph{Cowling  
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally letting $Q\to\infty$ in Eq. \ref{lin_poisson}).  
The growth rate increases when self-gravity is included in the linear
response, since self-gravity is usually destabilizing. However, 
$\gamma$ and $E_m$ are very similar, indicating that the instability
in the self-gravitating calculation is fundamentally still the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{
    Growth rates of MRI modes in an isothermal self-gravitating
    disk with $Q=0.2$ ($Q_\mathrm{2D}=0.72$) at fixed $\beta=100$, 
    for a range of midplane Elsasser numbers. The resistivity is
    uniform. The colorbar measures the importance of self-gravity by $\tau$. 
    \label{gravity_energy_resis}}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_resis_sg}
  \caption{Example of a resistive MRI mode with significant
    gravitational potential perturbation. The disk is isothermal. 
    The vertical line
    indicates $\Lambda=1$.  
    \label{mri_massive_resis}}
\end{figure}

@@

 However, a strong field and/or large resistivity increases the MRI
 vertical wavelength. When the vertical scale of the MRI becomes
 comparable to the disk thickness, i.e. $\lambda\sim H$, the
 perturbed mass across the disk height can contribute to a net potential
 perturbation. We therefore expect a necessary condition for
 self-gravity to affect the MRI is for the latter to be weak. 
%we are not considering GI yet

### 2014-04-05 paper/results_azi.tex
first draft, spell-checked

\section{Effect of an azimuthal field}\label{result3}
In this section we use the setup of case 2 described in
\S\ref{linear}, and examine the effect of an azimuthal field so that
$B_y\neq 0$,   
parameterized by $\epsilon \equiv B_y/B_z$. However, we continute to
use $B_z$ for normalizations and $\beta$ is associated with the
vertical Alfven speed. We also extend the previous calculations to the
full disk $z\in[-Z_s,Z_s]$,  which allows us to compare the effect of
self-gravity on MRI modes with different symmetries across the
midplane. We use an isothermal disk throughout.

\subsection{Ideal disks with MRI} 
We consider disk with $Q=0.2$ ($Q_\mathrm{2D}=~0.72$) and
$\beta=10$ in the limit of ideal MHD ($\Lambda_0=100$). Gravitational
instability is not expected because Fig. \ref{compare_growth3} shows
that even for $Q=0.18$, GI is surpressed for $\beta \lesssim 15$.

Fig. \ref{compare_growth3_tilted} show MRI growth rates for
$B_y/B_z=0,\,1,\,2$ and $3$. We divide the modes into two catagories
depending on the extremum of magnetic energy at the midplane. 
The top panel are modes where $E_m$ has a local minimum at $z=0$ and 
the bottom panel are modes where $E_m$ has a local maximum at $z=0$. 
The latter set of modes were excluded in the previous sections by
midplane boundary conditions. 
We also plot growth rates computed in the Cowling
approximation. As expected,  $\avg{E_g}<\avg{E_m}$, so none of the
modes are energetically dominated by self-gravity.

@@

Consider first modes in the top panel of 
Fig. \ref{compare_growth3_tilted}. As with previous results,  
self-gravity destabilizes modes with  $k_xH\gtrsim O(1)$. Consequently, the
cut-off wavenumber is larger when SG is included. 
Destablization is most effective for purely vertical fields: with
$\epsilon=0,\, k_xH\simeq 1.4$, SG increases the growth rate by $\sim
30\%$. For $B_y=0$ we find the density perturbation $W(z)$ is an even
function. Although these modes are fundamentally magnetic, this is consistent with
\cite{goldreich65a}, who showed that SG can only destabilize 
symmetric density perturbations.  
With increasing $B_y$, we find $W$ deviates from an even
function. 
Together with the increased total magnetic pressure with 
$B_y$ (since $B_z$ is fixed), destabilization by SG weakens. 
Thus, the Cowling approximation becomes increasingly good with stronger
$B_y$ for these modes.

@@

Compressibilty is enhanced by an azimuthal field, which is 
stabilizing for the MRI \citep{kim00}. %because...?
This effect is significant for $\epsilon=3$ because the azimuthal
Alfven speed is sonic.      
%maybe blaes 
Fig. \ref{result_tilted} indicates that self-gravity further enhances
compressibility, and therefore stabilization. We suspect this is
overwhelmed by the destabilization effect of SG, because the density
perturbation has an anti-symmetric component.

@@

Notice also in Fig. \ref{compare_growth3_tilted_resis} the disappearence of
magnetic modes between $0.8\lesssim k_xH\lesssim 1.5$ as $B_y$ is
increased. For $B_y=2B_z$, MRI and GI are again independent, but
because they operate at distinct radial scales. This implies that
perturbations unstable to GI cannot develop MRI.

>>>>

\section{Effect of an azimuthal field}\label{result3}
In this section we use the setup of case 2 described in
\S\ref{linear}, and examine the effect of an azimuthal field so that
$B_y\neq 0$,   
parametrized by $\epsilon \equiv B_y/B_z$. However, we continue to
use $B_z$ for normalizations and $\beta$ is associated with the
vertical Alfven speed. We also extend the previous calculations to the
full disk $z\in[-Z_s,Z_s]$,  which allows us to compare the effect of
self-gravity on MRI modes with different symmetries across the
midplane. We use an isothermal disk throughout.

\subsection{Ideal disks with MRI} 
We consider disk with $Q=0.2$ ($Q_\mathrm{2D}=~0.72$) and
$\beta=10$ in the limit of ideal MHD ($\Lambda_0=100$). Gravitational
instability is not expected because Fig. \ref{compare_growth3} shows
that even for $Q=0.18$, GI is suppressed for $\beta \lesssim 15$.

Fig. \ref{compare_growth3_tilted} show MRI growth rates for
$B_y/B_z=0,\,1,\,2$ and $3$. We divide the modes into two categories
depending on the extremum of magnetic energy at the midplane. 
The top panel are modes where $E_m$ has a local minimum at $z=0$ and 
the bottom panel are modes where $E_m$ has a local maximum at $z=0$. 
The latter set of modes were excluded in the previous sections by
midplane boundary conditions. 
We also plot growth rates computed in the Cowling
approximation. As expected,  $\avg{E_g}<\avg{E_m}$, so none of the
modes are energetically dominated by self-gravity.

@@

Consider first modes in the top panel of 
Fig. \ref{compare_growth3_tilted}. As with previous results,  
self-gravity destabilizes modes with  $k_xH\gtrsim O(1)$. Consequently, the
cut-off wavenumber is larger when SG is included. 
Destabilization is most effective for purely vertical fields: with
$\epsilon=0,\, k_xH\simeq 1.4$, SG increases the growth rate by $\sim
30\%$. For $B_y=0$ we find the density perturbation $W(z)$ is an even
function. Although these modes are fundamentally magnetic, this is consistent with
\cite{goldreich65a}, who showed that SG can only destabilize 
symmetric density perturbations.  
With increasing $B_y$, we find $W$ deviates from an even
function. 
Together with the increased total magnetic pressure with 
$B_y$ (since $B_z$ is fixed), destabilization by SG weakens. 
Thus, the Cowling approximation becomes increasingly good with stronger
$B_y$ for these modes.

@@

Compressibility is enhanced by an azimuthal field, which is 
stabilizing for the MRI \citep{kim00}. %because...?
This effect is significant for $\epsilon=3$ because the azimuthal
Alfven speed is sonic.      
%maybe blaes 
Fig. \ref{result_tilted} indicates that self-gravity further enhances
compressibility, and therefore stabilization. We suspect this is
overwhelmed by the destabilization effect of SG, because the density
perturbation has an anti-symmetric component.

@@

Notice also in Fig. \ref{compare_growth3_tilted_resis} the disappearance of
magnetic modes between $0.8\lesssim k_xH\lesssim 1.5$ as $B_y$ is
increased. For $B_y=2B_z$, MRI and GI are again independent, but
because they operate at distinct radial scales. This implies that
perturbations unstable to GI cannot develop MRI.

### 2014-04-05 paper/results_gi.tex
first draft, spell-checked

\subsection{Co-existence of MRI and GI} 
Fig. \ref{compare_growth3} show growth rates for modes with $k_xH=1$
as a function of $\beta$ in disks with $Q=0.18$, $Q=0.14$ and
$Q=0.12$.   
All three cases display distinct GI modes (red/brown branch). The GI 
growth rates are $\gamma\simeq 0.25\Omega,\,0.6\Omega,\,0.8\Omega$ for
$Q=0.18,\,0.14,\,0.12$, respecitvely. GI is stabilized by magnetic
pressure for sufficiently small $\beta$. The critical
field strength for stabilizating GI increases with increasing
self-gravity, consistent with \cite{nakamura83}.   
For $Q=0.18$, GI is stabilized for $\beta \lesssim 15$. Nevertheless, 
the MRI branch for $\beta < 15$ becomes self-gravitating, so that
density perturbations still grow, even though GI does not
formally operate.

@@

Thus, our results do not rigorously prove that the GI and MRI branches
do not intersect.  Nevertheless, the continous variation of $\tau$
strongly suggest that unstable modes can transition smoothly from 
MRI to GI and vice versa, especially at low $\beta$.     
%no unique transition

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth3_Q0d12_B20.ps}  
  \caption{Growth rates of unstable modes in the massive isothermal
    disk with $Q=0.12$ and $\beta=20$, as a function of the horizontal
    wavenumber $k_x$. The colorbar measures the importance of
    self-gravity by $\tau$. The solid line corresponds to MRI modes in the
    Cowling approximation. The dashed line corresponds to pure GI
    modes, obtained by including a high resisitivity in the full
    problem. 
    \label{compare_growth3_Q01d2}}
\end{figure}

>>>>

\subsection{Co-existence of MRI and GI} 
Fig. \ref{compare_growth3} show growth rates for modes with $k_xH=1$
as a function of $\beta$ in disks with $Q=0.18$, $Q=0.14$ and
$Q=0.12$.   
All three cases display distinct GI modes (red/brown branch). The GI 
growth rates are $\gamma\simeq 0.25\Omega,\,0.6\Omega,\,0.8\Omega$ for
$Q=0.18,\,0.14,\,0.12$, respectively. GI is stabilized by magnetic
pressure for sufficiently small $\beta$. The critical
field strength for stabilizing GI increases with increasing
self-gravity, consistent with \cite{nakamura83}.   
For $Q=0.18$, GI is stabilized for $\beta \lesssim 15$. Nevertheless, 
the MRI branch for $\beta < 15$ becomes self-gravitating, so that
density perturbations still grow, even though GI does not
formally operate.

@@

Thus, our results do not rigorously prove that the GI and MRI branches
do not intersect.  Nevertheless, the continuous variation of $\tau$
strongly suggest that unstable modes can transition smoothly from 
MRI to GI and vice versa, especially at low $\beta$.     
%no unique transition

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth3_Q0d12_B20.ps}  
  \caption{Growth rates of unstable modes in the massive isothermal
    disk with $Q=0.12$ and $\beta=20$, as a function of the horizontal
    wavenumber $k_x$. The colorbar measures the importance of
    self-gravity by $\tau$. The solid line corresponds to MRI modes in the
    Cowling approximation. The dashed line corresponds to pure GI
    modes, obtained by including a high resistivity in the full
    problem. 
    \label{compare_growth3_Q01d2}}
\end{figure}

### 2014-04-05 paper/setup.tex
first draft, spell-checked

The total potential is $\Phi = \Phi_\mathrm{ext} + \Phi_d$, where
\begin{align}
  \Phi_\mathrm{ext}(x,z) = -\Omega_0 S_0 x^2 +
  \frac{1}{2}\Omega_{z0}^2z^2 
\end{align}
is the effective external potential (central plus centrifugal) in the
shearing box approximation, where $S_0\equiv ~S(r_0,0)$ and
$\Omega_{z0}\equiv\Omega_z(r_0,0)$; 
and the gas potential $\Phi_d$ statisfies Poisson's equation
\begin{align}
  \nabla^2\Phi_d = 4\pi G \rho, 
\end{align}
where $G$ is the gravitational constant. For clarity, hereafter we
drop the subscript $0$ on the frequencies.

\subsection{Equilibrium disk} 
The unperturbed disk is steady and described by
$\rho=~\rho(z)$, $\bm{B} = B_z\hat{\bm{z}} + B_y\hat{\bm{y}}$ where
$B_{y,z}$ are constants and the toroidal field strength is
$B_y=\epsilon B_z$. The equilibrium 
velocity field is $\bm{v} = -Sx\hat{\bm{y}}$. We consider Keplerian
disks so that $S = 3\Omega/2$ and the epicyclic frequency 
$\kappa\equiv\sqrt{2\Omega(2\Omega-S)}=\Omega=\Omega_z$.  
We assume a thin disk and neglect the radial component of the
self-gravitational force in the unperturbed disk.

>>>>

The total potential is $\Phi = \Phi_\mathrm{ext} + \Phi_d$, where
\begin{align}
  \Phi_\mathrm{ext}(x,z) = -\Omega_0 S_0 x^2 +
  \frac{1}{2}\Omega_{z0}^2z^2 
\end{align}
is the effective external potential (central plus centrifugal) in the
shearing box approximation, where $S_0\equiv ~S(r_0,0)$ and
$\Omega_{z0}\equiv\Omega_z(r_0,0)$; 
and the gas potential $\Phi_d$ satisfies Poisson's equation
\begin{align}
  \nabla^2\Phi_d = 4\pi G \rho, 
\end{align}
where $G$ is the gravitational constant. For clarity, hereafter we
drop the subscript $0$ on the frequencies.

\subsection{Equilibrium disk} 
The unperturbed disk is steady and described by
$\rho=~\rho(z)$, $\bm{B} = B_z\hat{\bm{z}} + B_y\hat{\bm{y}}$ where
$B_{y,z}$ are constants and the toroidal field strength is
$B_y=\epsilon B_z$. The equilibrium 
velocity field is $\bm{v} = -Sx\hat{\bm{y}}$. We consider Keplerian
disks so that $S = 3\Omega/2$ and the epicycle frequency 
$\kappa\equiv\sqrt{2\Omega(2\Omega-S)}=\Omega=\Omega_z$.  
We assume a thin disk and neglect the radial component of the
self-gravitational force in the unperturbed disk.
